Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-22 paper/paper.tex
Minor comments

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that logical
relations-based arguments are not available.  \scw{Well, actually, we did use
  a step-indexed relation for Trellys\ldots} But this is not the case. Recent
authors have developed tour-de-force mechanizations for the metatheory of
modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have used logical-relations-based reasoning as part of their
arguments. However, at the same time, these definitions have been overshadowed
by their magnitude of the surrounding developments. Because these proofs show
diverse results about real systems and algorithms, these developments range in
size from 20,000 to 400,000 lines of code, and the techniques are not
accessible to a casual reader.

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, the but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. \scw{Say that we don't have impredicativity, but neither does anyone
  else?}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.

>>>>

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that logical
relations-based arguments are not available.  \scw{Well, actually, we did use
  a step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys' logical fragment
  is simple type + identity so I don't think that counts..} But this is not the case. Recent
authors have developed tour-de-force mechanizations for the metatheory of
modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have used logical-relations-based reasoning as part of their
arguments. However, at the same time, these definitions have been overshadowed
by their magnitude of the surrounding developments. Because these proofs show
diverse results about real systems and algorithms, these developments range in
size from 20,000 to 400,000 lines of code, and the techniques are not
accessible to a casual reader.

Our proof, designed to capture the essence of this proof technique in this
setting, requires less than 1000 lines of code. We have achieved this
significant reduction through a number of means: the careful selection of the
features that we include in the object type type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, the but we do include propositionality
equality and booleans in order to capture the challenges presented by indexed
types and dependent pattern matching. We do not show the decidability of type
checking, nor do we develop a PER semantics, but we do demonstrate how our
consistency proof can be extended (at a moderate cost of 500 lines of code) to
show the existence of $\beta\eta$-normal form for well-typed open \emph{and}
closed terms. We include a full predicative universe hiercharchy and
type-level computation, to demonstrate the logical strength of the
approach. \scw{Say that we don't have impredicativity, but neither does anyone
  else?}\yl{Nobody else but Barras. I'll find the citation}
%%
%% SCW: save specific comparisons with other systems for later
%%
% The big difference in the size of the developments does not
% necessarily imply that our proof technique leads to a more concise
% proof due to the differing expressiveness of the languages and the
% different metatheoretic results being established. For
% example, \citet{anand2014towards} mechanizes the metatheory of
% Nuprl~\citep{constable1986implementing}, which, on top of all the
% features that \lang{} supports, includes W-types and partial types and
% requires a PER semantics to model its extensional typed
% equality. The object language from \citet{decagda} does not support
% identity types and only has one predicative universe, but additionally
% supports $\Sigma$ types. However, for a researcher who wishes to learn
% the underlying techniques for mechanized logical relation for
% dependent type theory, a small development like ours is much easier to
% navigate and understand. Furthermore, we show how our simple
% consistency proof can be easily extended to show the existence of
% $\beta\eta$-normal form for well-typed open \emph{and} closed terms,
% giving us a metatheoretic result almost as strong as the one from
% \citet{decagda} at the moderate cost of around 500 lines of extra code.
